The Spatial and Temporal Domains of Modern Ecology

### 2016-07-07 paper/SI.tex
Commit of all changes (analysis, text) up to 6 July, 2016 (just before incorporating Erle's edits).

\section*{Supplementary Information}% Make the title.
\subsection*{Methods}% Make the title.
\vspace{-4pt}
%\maketitle 
\subsubsection*{Paper selection}
\vspace{-10pt}
We used the 2012 Web of Science impact factors to select the highest ranked ecology-themed journals that published studies with an observational component, excluding journals devoted to reviews, meta-analyses, laboratory, cellular, or experimental studies.  To select a representative sample of recent ecology studies, we downloaded the metadata for all papers published in the selected journals (Table S1) between 2004 and 2014. Our study involved six different observers (those reviewing the papers to extract the observational scales), each of whom was given a randomly selected batch of 500 titles.  A separate set of 20 randomly selected papers was also given to each observer, which formed the basis for calibrating observers' interpretation and extraction of scale-related information, as well as estimating between observer variance.

@@

Similarly, we did not try to quantify the extent to which values of individual observations (e.g. a single plot) of an ecological feature might be correlated with spatially or temporally adjacent values.  Autocorrelation is common to most ecological phenomena \cite{legendre_spatial_1993,collins_method_2000-1}, and it means that an observation is descriptive of the properties within a broader area or time around the point at which it is taken. For example, soil properties are typically measured by analysis of soil cores collected using an augur. The two-dimensional resolution of this measurement is equal to the area of the augur, but the measured soil properties are likely to be very similar, if not identical, to those found in the soil within a radius of at least a few to several hundreds of meters around the core location. However, the distance and strength of this correlation can vary substantially according to the ecological features being observed, and such correlation lengths were reported in almost no studies. Attempting to estimate these lengths, and add them to actual observations scales, would have added further uncertainty to estimates. Moreover, this concern only applies to spatially and temporally discontinuous (or discontiguous) observations, and since many of observations are collected using techniques that permit continuous measurements (e.g. weather data over time or satellite images over space), we elected to use a consistent standard for estimating dimensions.

In assessing spatial scales, our analysis only considered evaluated the Cartesian plane. We did not consider the z, or depth, dimension, which for certain aspects of ecology, this measure is of greater value (e.g. depth profiles in marine ecology), although in some cases (paleoecology) values extracted from the z-dimension provided a temporal measure that was used to calculated both the interval and the duration of the observation.

@@

  \Hitem \emph{\textbf{What about studies that sample individual organisms?}} If the study is making a total count of all organisms (let's say a mammal species) within a fixed plot size, or even a variable plot size from which an average plot size (and thus sampling resolution) can be estimated, then follow the procedures described in Q16-17. However, if the individual animals are the unit of measure (either because a sub-sample of them is being made within a defined plot, or because the observation is not contingent on being located within a plot (maybe a blood sample or body weight is being recorded, for example), then simply estimate two-dimensional area occupied by the animal, the number of animals sampled, and then calculate area from that. Occasionally individual animals might be recorded, but within the context of some natural feature, such as a nesting site where the survival of individual chicks is the measurement of interest \cite{roche_relative_2008-1}. In this case, an estimate of the nest area provides the sampling resolution. Another possible case is where individual animals are being tracked using radio or GPS collars. In these cases, use the estimate of the collared/tagged animal's body area, and then treat the number of reported locational observations as the number of spatial replicates, and then use those two values to calculate sampled area. The total GPS points collected from all individuals can be used if we can safely assume that each individual sampled did not cover the same exact location.  If the number of GPS points is not given, the number of fixes can be estimated from the duration during which individuals were collared and the recording interval.

@@

\subsubsection*{Calibration and consistency}
\vspace{-10pt}
Most studies did not explicitly report values for all four of the assessed scales, and thus interpretation and judgement had to be applied to develop reasonable estimates for their values. The FAQ provided the protocol we followed, and was initially developed following a consultative process prior to the commencement of review. We conducted an iterative process of calibration to ensure consistency and reliability of estimates. First, we used the calibration set to calculate between-observer variability with respect to paper selection/rejection and the estimation of scales. Based on this, the lead author reviewed individual records in each observer's calibration set, flagged values were the estimation procedure departed from the protocol, and returned these to observers for re-estimation, without providing an estimate of the actual value.  Instead, the relevant section of the protocol was highlighted, and further explanation and clarifying discussion undertaken as needed. Protocol language was adjusted for clarity during this process, and new items added to cover circumstances that had not been addressed by the initial version. The variability measures (see Results) were re-calculated after each iteration.

@@

\subsubsection*{Estimating observational density}
\vspace{-10pt}
In presenting results (Fig. 1 and 2 in main text), we log-transformed (base 10) the observational scales because of the large range in values.  To reveal the densities of observations within two dimensions (Fig. 2 main text), we used the splancs package \cite{rowlingson_splancs:_1993} of R \cite{r_development_core_team_r:_2011} to calculate a kernel density estimate of the log-transformed values, using a bandwidth of 1 onto a 0.1 resolution image to provide a smoothed result that served to more effectively highlight domains in which ecological observations are concentrated. Bandwidths of varying resolutions were tested (see Supplementary Results section).

\subsubsection*{Sensitivity of results to observer and estimation error}
\vspace{-10pt}
To test the impact of between-observer inconsistencies on results, as well as the possibility of error by individual observers made in estimating the scales of actual observations scale, we conducted a sensitivity analysis. For each of the four dimensions, we performed a 1000-iteration resample of values, adding to each record a random amount of error at each iteration. The error values were constrained to fall within a maximum percentage of the record's value, which was determined by the coefficient of variation between observers' estimates for the calibration set.  We then combined the set of 1000 perturbed observations, log-transformed them, recalculated the kernel density estimates for the comparisons of the two resolution and extent dimensions (i.e. Fig. 2A and B in main text), and assessed the degree to which the domains of observational concentration differed from those in our main analysis (Fig. S2).

@@

We assessed variability and consistency between observers along multiple dimensions. First, we assessed how reliablly observers agreed with respect to selecting or rejecting papers for review, using the R's Agreement package \cite{yu_agreement:_2012} to calculate Fleiss' kappa statistic \cite{fleiss_measuring_1971}, which was 0.72 (z = 12.5, p$<$0.000), indicating substantial and significant agreement between observers \cite{landis_measurement_1977}.

@@

\subsubsection*{Sensitivity analysis}
We used the maximum uncertainty values for each dimension from the inter-observer variability analysis (Table S2) to determine the bounds of the random perturbations applied to each record in the sensitivity analysis. Kernel density estimates applied to the merged results of this analysis for the contrasts between spatial resolution and temporal intervals, and to spatial extent and temporal duration, suggest that the primary features observed in Figures 2A and 2B (main text) are largely unaltered by the addition of uncertainty to the estimates. The largest difference is in the greater density of second to sub-second duration observations across all spatial extents.

\begin{figure}[!ht]
%\begin{wrapfigure}{c}{1\textwidth}
\includegraphics[width=1\textwidth]{figures/res_v_extent_pbstrap2.pdf}
\vspace{-0.15 cm}
\caption{Two-dimensional kernel density estimates of observational densities within the domains defined by sampling interval and spatial resolution (left) and temporal duration and spatial extent (right). Density estimates were applied to the log-transformed values of each observational dimension after being randomly perturbed within uncertainty bounds drawn from an assessment of between-observer variability. Density estimates are rescaled to represent percentages. }
\label{afoto1}
\end{figure}

Figure S2 indicates the effect that varying bandwidth has on the appearance of the kernel density estimates, as estimated for the same two contrasts shown in Figure S1. the smallest bandwidth (0.4) tested (Fig. S2 top) shows that the primary observational concentrations described in the main text are evident but harder to discern.  Most difficult to discern is the oblong concentration of observations that in Fig. 1A (main text) is bounded on the lower right at monthly to yearly intervals and 100-1000 m$^2$ resolutions and on the upper left by near-daily to monthly observations and 0.1-10 ha resolutions. With the smaller these appears as two separate patches (Fig. S2 top left), but with 0.7 bandwidth applied become coherent (Fig. S3 middle left).

\begin{figure}[!ht]
%\begin{wrapfigure}{c}{1\textwidth}
\includegraphics[width=0.8\textwidth]{figures/res_v_extent_ksize.pdf}
\vspace{10 pt}
\caption{Two-dimensional kernel density estimates of observational densities within the domains defined by sampling interval and spatial resolution (left column) and temporal duration and spatial extent (right column), applied to log-transformed values of each observational dimension. Rows indicate the effects of selecting different bandwidths: 0.4 (top row); 0.7 (middle row); 1 (bottom row). }
\label{afoto1}
\end{figure}

>>>>

\section*{Supplementary Information}% Make the title.
\subsection*{Methods}% Make the title.
\vspace{-4pt}
%\maketitle 
\subsubsection*{Paper selection}
\vspace{-10pt}
We used the 2012 Web of Science impact factors to select the highest ranked ecology-themed journals that published studies with an observational component, excluding journals devoted to reviews, meta-analyses, laboratory, cellular, or experimental studies.  To select a representative sample of recent ecology studies, we downloaded the metadata for all papers published in the selected journals (Table S1) between 2004 and 2014. Our study involved six different observers (those reviewing the papers to extract the observational scales), each of whom was given a randomly selected batch of 500 titles.  A separate set of 20 papers was also randomly selected, and this set was given to all observers to review, in order to 1) calibrate the interpretations and extraction of scale-related information between observers, and 2) to estimate between-observer variance.

@@

Similarly, we did not try to quantify the extent to which values of individual observations (e.g. a single plot) of an ecological feature might be correlated with spatially or temporally adjacent values.  Autocorrelation is common to most ecological phenomena \cite{legendre_spatial_1993,collins_method_2000-1}, and it means that an observation can be representative of the properties within a broader area or time around the point at which it is taken. For example, soil properties are typically measured by analysis of soil cores collected using an augur. The two-dimensional resolution of this measurement is equal to the area of the augur, but the measured soil properties are likely to be very similar, if not identical, to those found in the soil within a radius of at least a few to several hundreds of meters around the core location. However, the distance and strength of this correlation can vary substantially according to the ecological features being observed, and such correlation lengths were reported in almost no studies. Attempting to estimate these lengths, and add them to actual observations scales, would have added further uncertainty to estimates. Moreover, this concern only applies to spatially and temporally discontinuous (or discontiguous) observations, and since many of observations are collected using techniques that permit continuous measurements (e.g. weather data over time or satellite images over space), we elected to use a consistent standard for estimating dimensions.

In assessing spatial scales, our analysis only considered the Cartesian plane. We did not calculate the z, or depth, dimension, although this dimension is of greater importance for certain sub-disciplines of ecology (e.g. depth profiles in marine ecology). In some cases (primarily paleoecological studies), values extracted from the z-dimension provided temporal information that was used to calculate both the interval and the duration of the observation.

@@

  \Hitem \emph{\textbf{What about studies that sample individual organisms?}} If the study is making a total count of all organisms (let's say a mammal species) within a fixed plot size, or even a variable plot size from which an average plot size (and thus sampling resolution) can be estimated, then follow the procedures described in Q16-17. However, if the individual animals are the unit of measure (either because a sub-sample of them is being made within a defined plot, or because the observation is not contingent on being located within a plot (maybe a blood sample or body weight is being recorded, for example), then simply estimate two-dimensional area occupied by the animal, the number of animals sampled, and then calculate area from that. Occasionally individual animals might be recorded, but within the context of some natural feature, such as a nesting site where the survival of individual chicks is the measurement of interest \cite{roche_relative_2008}. In this case, an estimate of the nest area provides the sampling resolution. Another possible case is where individual animals are being tracked using radio or GPS collars. In these cases, use the estimate of the collared/tagged animal's body area, and then treat the number of reported locational observations as the number of spatial replicates, and then use those two values to calculate sampled area. The total GPS points collected from all individuals can be used if we can safely assume that each individual sampled did not cover the same exact location.  If the number of GPS points is not given, the number of fixes can be estimated from the duration during which individuals were collared and the recording interval.

@@

\subsubsection*{Calibration and consistency}
\vspace{-10pt}
Most studies did not explicitly report values for all four of the assessed scales, and thus interpretation and judgement had to be applied to develop reasonable estimates for their values. The FAQ provided the protocol we followed, and was initially developed following consultation between observers prior to the commencement of review. We conducted an iterative process of calibration to ensure consistency and reliability of estimates. First, we used the calibration set to calculate between-observer variability with respect to paper selection/rejection and the estimation of scales.  Based on this, the lead author reviewed individual records in each observer's calibration set, flagged values where the estimation procedure departed from the protocol, and returned these to observers for re-estimation, without providing an estimate of the actual value. Instead, the relevant section of the protocol was highlighted, and further explanation and clarifying discussion undertaken as needed. Protocol language was adjusted for clarity during this process, and new items added to cover circumstances that had not been addressed by the initial version. The variability measures (see Results) were re-calculated after each iteration.

@@

\subsubsection*{Accounting for estimation uncertainty}
\vspace{-10pt}
Two major sources of uncertainty affected our estimation of observational scales: 1) unclear documentation of observational scales in the reviewed studies, 2) variation between observers in estimating observational scales. We estimated and accounted for these uncertainties in several ways. First, we estimated the degree of between-observer variability based on each observer's results from reviewing the 20-paper calibration set. We calculated how well observers agreed regarding paper inclusion/exclusion, how many extractable observations there were per included paper, and what the coefficient of variation was across all observers' estimates of scale. We also recorded when observations were reported in a study with an unclear or missing scale value, which we used to assess a dimension-specific estimate of uncertainty.

We used the between-observer coefficients of variation for each dimension as the basis for randomly perturbing the scale values for each of the 371 observations over 1000 iterations. For each observation at each iteration, we perturbed its scale value in each dimension by randomly selecting a percentage value p that fell between 100 + y and 100 - y, where y was the between-observer coefficient of variation (expressed as a percentage) for a given observational dimension, and multiplying that observation's scale value by that proportion. This perturbed set of observations provided a basis for estimating uncertainty within our extracted set of observational scales.

\subsubsection*{Scale metrics}
\vspace{-10pt}
In presenting results (Fig. 1 and 2 in main text), we log-transformed (base 10) the observational scales within each of the 1000 perturbed ensemble members in order to account for the large range in scale values. We calculated histograms for each observational dimension from each of the log-transformed ensemble members, calculating the mean percentage density estimate for each bin across all 1000 histograms, as well as the upper and lower 2.5th percentile values for each bin (Fig. 1 main text).

To reveal the densities of observations within two dimensions (Fig. 2 main text), we used the splancs package \cite{rowlingson_splancs:_1993} of R \cite{r_development_core_team_r:_2011} to calculate a kernel density estimate of the log-transformed values across all ensemble members, using a bandwidth of 1 onto a 0.1 resolution image to provide a smoothed result that served to more effectively highlight domains in which ecological observations are concentrated. Bandwidths of varying resolutions were tested on kernel density estimates were tested on kernel density estimates of sampling interval versus plot resolution (see Supplementary Results section).

@@

We assessed variability and consistency between observers along multiple dimensions. First, we assessed how reliably observers agreed with respect to selecting or rejecting papers for review, using the R's Agreement package \cite{yu_agreement:_2012} to calculate Fleiss' kappa statistic \cite{fleiss_measuring_1971}, which was 0.72 (z = 12.5, p$<$0.000), indicating substantial and significant agreement between observers \cite{landis_measurement_1977}.

@@

\subsubsection*{Between-observer coefficient of variation}
We used the maximum uncertainty values for each dimension from the inter-observer variability analysis (Table S2) to determine the bounds of the random perturbations applied to each record over the 1000 iteration resample.

%\begin{figure}[!ht]
%%\begin{wrapfigure}{c}{1\textwidth}
%\includegraphics[width=1\textwidth]{figures/res_v_extent_pbstrap2.pdf}
%\vspace{-0.15 cm}
%\caption{Two-dimensional kernel density estimates of observational densities within the domains defined by sampling interval and spatial resolution (left) and temporal duration and spatial extent (right). Density estimates were applied to the log-transformed values of each observational dimension after being randomly perturbed within uncertainty bounds drawn from an assessment of between-observer variability. Density estimates are rescaled to represent percentages. }
%\label{afoto1}
%\end{figure}
%
Figure S1 indicates the effect that varying bandwidth has on the appearance of the kernel density estimates. The smallest bandwidth (0.4) tested (Fig. S2 top) shows that the primary observational concentrations described in the main text are evident but harder to discern.  Most difficult to discern is the oblong concentration of observations that in Fig. 1A (main text) is bounded on the lower right at monthly to yearly intervals and 100-1000 m$^2$ resolutions and on the upper left by near-daily to monthly observations and 0.1-10 ha resolutions. With the smaller bandwidth this concentration appears as two separate patches (Fig. S2 top left), but with 0.7 bandwidth applied becomes coherent (Fig. S3 middle left).

\begin{figure}[!ht]
%\begin{wrapfigure}{c}{1\textwidth}
\includegraphics[width=0.9\textwidth]{figures/res_v_extent_ksize2.pdf}
\vspace{10 pt}
\caption{Two-dimensional kernel density estimates of observational densities within the domains defined by sampling interval and spatial resolution (left column) and temporal duration and spatial extent (right column), applied to log-transformed values of each observational dimension. Rows indicate the effects of selecting different bandwidths: 0.4 (top row); 0.7 (middle row); 1 (bottom row). }
\label{afoto1}
\end{figure}
